\begin{abstract}
\textbf{\textendash\textendash \textit{Background: }}Several decades have passed since the first experiment was carried out in Software Engineering (SE). Since then, there have been a significant upturn regarding the number of experiments undertaken in academy and lately in industry. However, some SE researchers consider that there is still concern about the activities to be performed before, during and after the experiment. \textbf{\textit{Aim: }}Identify and validate constraints surrounding the experimentation in SE, from the perspective of the signs of formality identified into the experimental process of tradicional disciplines. \textbf{\textit{Method: }}An exploratory study based on empirical mix methods was conducted, which involves the following stages: (1) identify the activities' complexity regarding the experimentation in SE; (2) learn about the experimental process performed in a traditional discipline and identify particular constraints within experimentation in SE through comparison; and (3) validate the presence of experimental SE constraints (identified in stage 2) in the rest of the ESE community. \textbf{\textit{Results: }}Conceptual and process models were obtained around the observation and participation process of the experimental activities in SE and other traditional discipline. Some constraints were observed in the SE experimentation compared with the experimental process of a traditional discipline, which are common into the ESE community. \textbf{\textit{Conclusiones: }}There are activities and concepts used in the experimentation of traditional disciplines similar to those used in the SE. However, there are additional activities and concepts that could be applied in SE in order to improve its experimental process.
\end{abstract}
%\textbf{\textendash\textendash \textit{Antecedentes: }}Han pasado varias décadas desde que se llevó a cabo el primer experimento en Ingeniería de Software (SE). Desde entonces, ha habido un importante repunte en el número de experimentos realizados en la academia y últimamente en la industria. No obstante, algunos investigadores en SE consideran que aún existe incertidumbre respecto a las actividades a realizar antes, durante y después de un experimento. \textbf{\textit{Objetivo: }}Identificar y validar la problemática en torno a la Experimentación en SE, tomando como referencia los indicios de formalidad identificados en el proceso de experimentación de las disciplinas tradicionales. \textbf{\textit{Metodología: }}Se realizó un estudio exploratorio basado en métodos empíricos mixtos para: (1) indagar sobre la problemática en torno a la experimentación en SE dentro de un grupo de investigación representativo en la comunidad de SE empírica (ESE), (2) aprender del proceso de experimentación en una disciplina tradicional e identificar la problemática particular de la Experimentación en SE, y (3) validar la generalidad de la problemática identificada en la experimentación en SE. \textbf{\textit{Resultados: }}Se obtuvieron modelos conceptuales y de proceso que permitieron entender de forma explícita la experimentación llevada a cabo en la SE, así como sus carencias respecto a una disciplina experimental tradicional y finalmente se comprobó que las carencias identificadas en la SE son generales. \textbf{\textit{Conclusiones: }} Existen actividades y conceptos utilizados en el proceso de experimentación de las disciplinas tradicionales que son semejantes a los utilizados en el proceso de experimentación es SE. Sin embargo, existen actividades y conceptos que podrían ser aplicados en la SE para mejorar su proceso de experimentación.