\section{Introduction}\label{sec-introduction}
%More than 50 years have passed since the first experiment was carried out into SE \cite[p.~7]{Wohlin2012-Experimentation} as mechanism to assess current methods and techniques of software development in order to improve the software product quality. Several experiments have been performed both in academy \cite{Sjoberg2005-surveyexperimentsESE} and industry \cite{dieste-2013-industry-experiments}. Nowadays it is difficult to find conferences or journals that does not include experimental studies. This situation prompted that SE researchers, interested in experimentation, had to acquire adequate training to take advantage of the different sources of information regarding the experimental process (e.g.: booklets, textbooks, guidelines, papers, among others.)\cite[p.~3-4]{Kitchenham-2015-Literature-Review-Book}. Therefore, the experimentation in SE appears to be a seated methodology.

%Experimentation in SE appear to be a probed methodology, taking into account the amount of years that have passed since the first experiment was carried out and the vast literature generated in this discipline \cite[p.~7]{Wohlin2012-Experimentation}. Nowadays it is difficult to find conferences or journals in which are not included experimental studies.

Experimentation is a well-established research method in SE nowadays. It is difficult to find conferences or journals in SE where experimental studies not be included. In fact, researchers in SE have started to perform an introspection of its discipline, as it happened in other traditional disciplines, such as Medicine \cite{Baumgardner-1997-statistical-issues-medicine}. For example, Kitchenham et al. \cite{Kitchenham2002-GuideLinesESE} point it out that: \textquotedblleft If researchers have difficulty in a discipline such as medicine, which has a rich history of empirical research, it is hardly surprising that software engineering researchers have problems\textquotedblright. On the other hand, inadequate procedures of statistical analyses \cite{Reyes-2009-Statistical-Errors-in-SE} and problems derived from researcher and publication bias \cite{Jorgensen-2016-Incorrects-Results-SEE} have already been identified. However, there remains important details within experimentation that apparently have not yet been fully suitable in SE and still represent a challenge.

An aspect which begins to attract attention nowadays, is the high cost of carrying out SE experiments in terms of difficulty, number of participants and time consuming; despite it has already been clearly identified in the past \cite{mian-200-computarized-infraestructure}. There are some incidents within the experimental environment of SE that apparently depict this fact, as shown below.

\begin{itemize}
  \item There is a growing trend in SE to carry out experiments \cite{Sjoberg2005-surveyexperimentsESE,Zendler2001,Dieste2011-CompAnalMetaWhenWich}, but the percentage of replications performed so far does not harmonize to such experimentation \cite{da-silva-2012-replication-on-empirical-studies}, which could minimize the validity of its findings.
  \item According to several authors \cite{Demagalhaes-2015-SMS-Replications,Gomez-2014-understanding-replication,Carver-2014-Replications-of-SE}, the lack of replications in SE is due primarily to there is uncertainty at the time of opting for one of the several existing mechanisms of replication, and that, none of them has been totally effective.
  \item There are basic guidelines to perform experiments and hence replications in SE \cite{Juristo2001,Wohlin2000,Wohlin2012-Experimentation,Basili-1986-ESE,Kitchenham2002-GuideLinesESE}, as well as to report it \cite{Jedlitschka2008,Kitchenham:2008b,Carver2010-GuidelinesReplication}; however, the variety of experimental protocols and report formats applied in practice far exceeds those methodological proposals \cite{Demagalhaes-2015-SMS-Replications}.
  \item There are groups of similar experimental studies (statistically speaking) applying different statistical techniques for their analysis, that in some cases incurre in methodological errors, which cause uncertainty at the time of opting for one of the several existing experiments' analysis techniques \cite{Reyes-2009-Statistical-Errors-in-SE}.
  \item Practitioners still have serious doubts about the validity of findings obtained from empirical research performed in SE \cite{Lo-2015-Practitioners-Perceive}, despite the vast amount of studies carried out in academia \cite{Sjoberg2005-surveyexperimentsESE} and lately in industry \cite{dieste-2013-industry-experiments}. 
\end{itemize}

%result in low trustworthiness of results reported in software engineering experiments.
%
%The lack or misuse of experimental design \cite{Dyba2006} increases the threats to validity around the experiments.
%
%The lack of communication between researchers \cite{Shull-2004-Knowledge-Sharing-Issues} generates problems around  sharing and storage of experimental information.
%
%The amount of experimental replication in SE is limited, mainly due to its complexity \cite{da-silva-2012-replication-on-empirical-studies}; hence, it's difficult to generalize the experimental results.
%
%The dearth of uniformity in the experimental reports \cite{Jedlitschka2008} makes difficult the information access.

% However, the current situation of the experimentation in SE is less satisfactory than it could be. Kitchenham et al. \cite{Kitchenham2002-GuideLinesESE}, for example, point it out that: \textquotedblleft If researchers have difficulty in a discipline such as medicine, which has a rich history of empirical research, it is hardly surprising that software engineering researchers have problems\textquotedblright.  

%However, in addition to this characteristics, they include the use of specialized vocabulary, and the decision making conflicting (on many occasions) with similar empirical studies, during the stages of design implementation and analysis, which causes that each process is completely different.

%Nowadays, there are uncertainties about how to perform experiments in terms of design \cite{Jorgensen-2016-Incorrects-Results-SEE} and execution \cite{mian-200-computarized-infraestructure}. According some studies the experimentation in SE has several deficiencies, for example: the possibility to follow different protocols to perform experiments, results in uncertainties regarding the correct option to take \cite{Wohlin2012-Experimentation}. The lack or misuse of experimental design \cite{Dyba2006} increases the threats to validity around the experiments. The lack of communication between researchers \cite{Shull-2004-Knowledge-Sharing-Issues} generates problems around  sharing and storage of experimental information. The amount of experimental replication in SE is limited, mainly due to its complexity \cite{da-silva-2012-replication-on-empirical-studies}; hence, it's difficult to generalize the experimental results. The dearth of uniformity in the experimental reports \cite{Jedlitschka2008} makes difficult the information access.

%Nowadays, there are uncertainties about how to perform experiments in terms of design \cite{Jorgensen-2016-Incorrects-Results-SEE} and execution \cite{mian-200-computarized-infraestructure}. According some studies the experimentation in SE has several methodological deficiencies, for example: the possibility to follow different protocols \cite{Juristo2001,Wohlin2000,Wohlin2012-Experimentation,Basili1985-ExperimentationSE,Kitchenham2002-GuideLinesESE,Jedlitschka2005-GuideLinesESE} to perform experiments results in uncertainties regarding the correct option to take.  Researcher and publication bias  \cite{Jorgensen-2016-Incorrects-Results-SEE} and inappropriate statistical procedures \cite{Reyes-2009-Statistical-Errors-in-SE}, such as the inappropriate use of techniques to formulate and evaluate statistical hypothesis, taking into account the meaningful conclusions obtained from it \cite{Miller-2004-statistical-significance};  

%Low trustworthiness of results reported in software engineering experiments mainly caused by researcher and publication bias, given typical statistical power and significance levels \cite{Jorgensen-2016-Incorrects-Results-SEE}.

%[OJO QUE ESTO AUMENTÉ AL PRINCIPIO, PERO DEBE QUEDAR COMO ANTES. ESTO PODRÍA SER UTILIZADO MÁS ABAJO EN DONDE HABLA DE LOS PROBLEMAS PUNTUALES: Nowadays, there are uncertainties about how to perform experiments in terms of avoid complex design with many statistical and unjustified tests \cite{Jorgensen-2016-Incorrects-Results-SEE} and execution \cite{Mendonca-2008-FIRE}.]

Such incidents within experimentation in SE aren't privative of this discipline. Other experimental disciplines, such as Medicine and Psychology, reported similar problems at some point \cite{Garcia-2012-Cienci-Abierta}. But now, experimentation in such disciplines is currently a very rigorous and uniform process \cite{stoker-2009-iLAP-bioinformatics,Jones-2007-FUGE,ko-2012-SMISB}.

%Esto se corrobora al observar de cerca la realización de experimentos en dichas áreas, donde parece que el conocimiento es mucho más compartido. Esta característica induce a pensar en que la definición del protocolo experimental en otras disciplinas es más formal.

%Cabe preguntarse entonces \textquestiondown En qué medida la diversidad de protocolos experimentales aplicados en la realización de un experimento desfavorece a la adopción de la experimentación en SE? y \textquestiondown Cuán madura es la experimentación en SE respecto a otras disciplinas experimentales tradicionales?. These questions motivated to carry out an exploratory study to research about the underlying incidents surrounding the ESE and inquire about their possible origins. This empirical study was based on mix methods, in order to identificar los problemas en un entorno específico donde se lleva cabo experimentación en SE, obtain insights (survey) of the most representative researchers into the Empirical Software Engineering Community about ESE incidents y finalmente, comparar en la práctica el proceso de ESE estudiado con el de una disciplina experimental tradicional.  
 
The question that arises then is:

\begin{framed}%
\textbf{RQ: What are the similarities and differences in implementing the experimental protocol between SE and other experimental disciplines?}
\end{framed}
 
%In order to perform the investigation to give an answer to the RQ, a mixed research method was applied. First, we conducted an ethnographical study in a representative SE research group, in order to observe specific details about its experimental protocol. Secondly, we aimed at extrapolating our findings to the entire SE community by means of a survey. Finally, we conducted a second ethnographical study in a non-SE (biotechnology) research group, in order to compare our findings
%OSCAR: CREO QUE DE ACUERDO A LO QUE HABLAMOS, EL PÁRRAFO ANTERIOR DEBERÍA IR EN LUGAR DEL QUE ESTA APARECIENDO, DADO QUE EN ESE ORDEN VAN APARECER LAS SECCIONES, DE LO CONTRARIO YA NO SÉ SI PRIMERO APARECE EL SURVEY Y LUEGO LA ETNOGRAFÍA DE BIOTECNOLOGÍA.

In order to answer the RQ, a mixed research method was applied. First, we conducted two ethnographical studies, in order to observe specific details about the experimental protocol applied in each case. Secondly, we aimed at extrapolating our findings to the entire SE community by means of a survey.   

Como resultado de la investigación encontramos que dentro de un grupo de investigación en SE no se sigue un protocolo experimental único, dado que cada experimentador aplica su propia terminología y criterio al realizar las actividades del proceso experimental que tiene a su cargo, lo que dificulta la comunicación de información dentro y fuera del grupo, e influye negativamente en la integración de información y en la realización de replicaciones. La comunidad de ESE al parecer esta consciente de esto, lo que indica que esta situación nos es privativa del grupo estudiado, sino de toda la comunidad. Sin embargo, a pesar de que en las disciplinas experimentales maduras (como la Biotecnología) existen los mismos problemas que en IS a nivel general, la diferencia se marca con la existencia de un protocolo experimental bien definido por área específica (e.g.: nutrición, patógenos de plantas, entre otras) que guía el proceso de experimentación; el cual promueve la uniformidad del trabajo, la fluidez en la comunicación entre los investigadores, dado que cada experimentador realiza tareas puntuales en el proceso de experimentación.

Este paper contribuye a la investigación en SE mediante la identificación de actividades experimentales de otras disciplinas que mejoran la formalización del proceso experimental en áreas específicas. Se precisa alcanzar la armonización de las tareas experimentales en SE, con el propósito de obtener protocolos específicos por cada área de investigación, tal que represente a la SE una solución más plausible, para de esa manera favorecer al intercambio de información en particular y a la experimentación en general. 

Lo que resta del artículo está estructurado como se indica a continuación. En la Sección \ref{sec-background}, se describen los antecedentes que motivaron la investigación. A continuación en la Sección \ref{sec-methology}, se detalla el método de trabajo aplicado en la investigación. La Sección \ref{sec-ESE-etnography} describe a detalle el estudio etnográfico sobre experimentación llevada a cabo en un grupo de investigación en SE. A paso seguido, en La Sección \ref{sec-survey} se describe el proceso de validación de la problemática en torno a la ESE identificada. En la Sección \ref{sec-bio-etnography} se presenta el estudio etnográfico realizado en áreas específicas de la Biotecnología. En la Sección \ref{sec-threats}, se describen las amenazas a la validez de la investigación y la forma en que se planteó su mitigación. Finalmente, una discusión en torno a las lecciones aprendidas y las conclusiones generales, es descrita en la Sección \ref{sec-discussion-conclusions}.